### 2023-03-28 paper-thesis/chapters/01-introduction.tex
More writing of the future work

\section*{Contributions}
This goal of this thesis is to investigate how well Statix is fit for the task of defining a dependently-typed language. The goal is not only to show that Statix can implement it, but also that the implementation is more concise than in a general purpose language. We will start by implementing the Calculus of Constructions~\cite{Coquand_Huet_1988}, a lambda calculus with dependent types, as this is one of the most simple dependently typed languages (chapter~\ref{chap:baselang}).

However, this language is not very practical to write actual complex programs in, so to improve this situation we will extend the language with typical features of dependently typed languages. We will do this in the following steps:
\begin{itemize}
	\item We show the language is easily extendable, by extending it with booleans among other features (Chapter~\ref{chap:bools})
	\item We show how to add inference of implicit arguments to the implementation (Chapter~\ref{ch:inference})
	\item We show how to add support for inductive datatypes to the implementation (Chapter~\ref{ch:datatypes})
	\item We show how to add support for universes to the implementation (Chapter~\ref{ch:universes})
\end{itemize}

>>>>

\section*{Contributions}
The primary contribution of this thesis is to investigate how well Statix is fit for the task of defining a dependently-typed language. The goal is not only to show that Statix can implement it, but also that the implementation is more concise than in a general purpose language. We will start by implementing the Calculus of Constructions~\cite{Coquand_Huet_1988}, a lambda calculus with dependent types, as this is one of the most simple dependently typed languages (chapter~\ref{chap:baselang}).

However, this language is not very practical to write actual complex programs in, so to improve this situation we will extend the language with typical features of dependently typed languages. We will provide the following additional contributions:
\begin{itemize}
	\item We show the language is easily extendable, by extending it with booleans among other features (Chapter~\ref{chap:bools})
	\item We show how to add inference of implicit arguments to the implementation (Chapter~\ref{ch:inference})
	\item We show how to add support for inductive datatypes to the implementation (Chapter~\ref{ch:datatypes})
	\item We show how to add support for universes to the implementation (Chapter~\ref{ch:universes})
\end{itemize}

### 2023-03-28 paper-thesis/chapters/98-relatedwork.tex
More writing of the future work

In chapter \ref{ch:comp-haskell} and \ref{ch:comp-lambdapi} we already compared our implementation with two different implementations. In this chapter we discuss remaining related work.
TODO TODO TODO

@@

The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope. Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types}~\cite{nameres} present an implementation of System F in Statix that does lazy substitutions, by using scopes as types. It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can be arbitrary terms.

@@

Construction Language Implementations~\cite{Rouvoet21} describes how to create a declarative specification of a language in Statix, and then to use a dependently typed language to integrate the specification of well-typing in the representation of the program that is being interpreted or transformed. The work from this thesis and ours could be combined, by using our dependently typed languages to do this verification.

@@

Below are some dependently typed languages, we discuss what make these languages unique and if these languages could be implemented in Statix:

@@

In chapter~\ref{ch:comp-haskell} we compared the implementation in Statix with an implementation in Haskell. The pi-forall language~\cite{pi_forall} is a language with a similar complexity to the language presented in this paper. In principle, the implementations are very similar. For example, the inference rules of pi-forall are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper. The primary difference is that they use a bidirectional type system~\cite{bidirectional}, whereas this paper uses Statix' unification.

\paragraph{Other}

>>>>

We now discuss some work that is related to the topics in this thesis. This thesis implements a dependently typed language in Statix. We compare this implementation with implementations of other languages in Statix, and with implementations of other dependently typed languages.

@@

The implementation in this paper requires performing substitutions in types immediately, as types don't have a corresponding scope to store substitutions. Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types}~\cite{nameres} present an implementation of System F in Statix that does lazy substitutions, by using scopes as types. Each scope represents a type, and scopes have edges to other scopes in case of quantifications and substitutions. Further research is needed to see if this approach could also apply to dependently typed languages, where types can contain arbitrary terms.

@@

Construction Language Implementations~\cite{Rouvoet21} describes how to create a declarative specification of a language in Statix, and then to use a dependently typed language to integrate the specification of well-typing in the representation of the program that is being interpreted or transformed. The work from this thesis and ours could be combined, by using our dependently typed language implemented in Statix, to do this verification, as another Statix DSL. This could make the presented approach integrate better with the ecosystem of Spoofax.

@@

Below are some more implementations dependently typed languages, we discuss what make these languages unique and if these languages could be implemented in Statix:

@@

In chapter~\ref{ch:comp-haskell} we compared the implementation in Statix with an implementation in Haskell. The pi-forall language~\cite{pi_forall} is a language with a similar complexity to the language presented in this thesis. In principle, the implementations are very similar. For example, the type checking rules of pi-forall are similar to the type checking rules presented in figure \ref{fig:type-check-rules} from this paper. The primary difference is that they use a bidirectional type system~\cite{bidirectional} to do some inference, whereas this paper uses Statix' unification.

\paragraph{Agda}

Agda is one of the most used dependently typed languages~\cite{agda}. It is implemented in Haskell, with full editor support for Emacs, VS Code, Atom and Vim. Interacting with the language is often done by giving commands to a language server, for example one can press \verb|Ctrl+C Ctrl+A| to do automatic proof search. These commands give feedback in a separate view in the editor.

Other dependently typed languages such as Lean and Coq have similar views to interact with the language. In general, when working with dependently typed languages there is a lot of these kinds of interactions between the IDE and the developer. Spoofax currently does not have support for defining shortcuts or adding these views to Eclipse, but there are no fundamental issues that would prevent this from being implemented.

### 2023-03-28 paper-thesis/chapters/99-conclusion.tex
More writing of the future work

\paragraph{RQ3: Can we add inference to the implementation?}
In chapter~\ref{ch:inference} we discuss how to add inference to the implementation. If we want to keep the implementation clean and concise, we need to compromise on how powerful the inference algorithm is, by defining an approximated version of first-order inference. We implemented this algorithm in a concise way.

@@

>>>>

\paragraph{RQ3: Can we add inference to the implementation?}
In chapter~\ref{ch:inference} we discuss how to add inference to the implementation. If we want to keep the implementation clean and concise, we need to compromise on how powerful the inference algorithm is, by defining an approximated version of first-order inference. We implemented this algorithm in a concise way. In chapter~\ref{ch:comp-lambdapi} we explore how we could define the language in a language supporting equational unification, adding this to Statix would allows for a more powerful inference algorithm to be cleanly implemented.

@@

\subsection*{Adding more features to the language}

In this thesis we created a language that already has a lot of features: Inductive datatypes, universes, and a basic inference algorithm. In chapter~\ref{chap:bools} we also specified the general steps that are needed to extend the language with new constructs. However, in comparison with languages such as Agda a lot of features are still missing, among which are:

\begin{itemize}
	\item Support for universe polymorphism~\cite[Universe Levels]{agda}, which allow functions and inductive data types to be polymorphic over a certain universe level. For example, this allows to define a \verb|List| data type that can be both a list of values and a list of types without defining it twice. We see no reason why this couldn't be cleanly implemented. 
	\item Another feature of Agda is implementing dependent pattern matching, as described in "Eliminating Pattern Matching without K"~\cite{patternmatching}. This is requires a desugaring stage, in which matches are desugared to eliminators, which should be implemented in Stratego before type checking. This should also be possible to implement, but it may be quite a bit of work.
	\item Inference in our language is currently implemented by explicitly specifying a placeholder \verb|_| where inference is required. In Agda we can specify an implicit argument, which is automatically inferred unless explicitly provided. We see no reason why this couldn't be cleanly implemented. 
	\item Instance Arguments are a special kind of implicit arguments, that are solved by the instance resolution algorithm of Agda~\cite[Instance Arguments]{agda}. They can be used similarly to type class constraints, in order to restrict the data types over which a term is polymorphic, for example requiring the terms to have an implementation of equality. The instance resolution algorithm should be possible to implementing by traversing the scope graph (since it is linear), and checking whether each let binding is a possible solution similar to the Agda implementation. 
	\item Agda has holes, which allows the user of the language to build proofs interactively. One of these interactions is proof search, when triggered Agda will try to find a term that fits the expected type of a hole. The interactive part of this was discussed in chapter~\ref{ch:relatedwork}, this is not yet supported by Spoofax. After support for this is added, a proof search algorithm could be implemented in Stratego, such as the structurally recursive unification algorithm described by McBride~\cite{proofsearch}.
	\item Agda allows function arguments to be marked as irrelevant, meaning that the value will be erased at runtime. An even stronger version of this feature is implementing a quantitive type system, such as in Idris~2~\cite{idris2}, which combines linear and dependent types. Implementing this in Spoofax should be possible, but it may be quite a bit of work as it is a complex type system.
	\item Agda has mixfix operators, which allows to define custom operators such as \verb|a and b|. This requires parsing to happen in multiple stages, where the first stage finds the mixfix operator definitions and the second stage parses expressions using them. This results in quite a complex parser, implementing these multiple stages in Spoofax using SDF3 is not possible without fundamental changes to the way that Spoofax works. One solution would be using an \emph{adaptive parsing algorithm}, which allows modification of the grammar during parsing. More research is needed to explore the possibilities of solving this problem.
	\item Modules
	\item Recursion
\end{itemize}
